% =========================================================================
% ACL-Format Paper: Generative vs Discriminative Text Classification
% Self-contained — compiles directly on Overleaf without external .sty files
% =========================================================================

\documentclass[11pt,a4paper,twocolumn]{article}

% --- Page geometry matching ACL format ---
\usepackage[margin=1in, columnsep=0.25in]{geometry}

% --- Typography ---
\usepackage{times}
\usepackage{microtype}
\usepackage[T1]{fontenc}

% --- Math ---
\usepackage{amsmath,amssymb}

% --- Tables ---
\usepackage{booktabs,multirow,array}

% --- Graphics ---
\usepackage{graphicx,float}

% --- Hyperlinks ---
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{url}

% --- Captions ---
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% --- Title formatting (ACL-style) ---
\usepackage{titling}
\setlength{\droptitle}{-1.5cm}
\pretitle{\begin{center}\Large\bfseries}
\posttitle{\par\end{center}\vskip 0.5em}
\preauthor{\begin{center}\normalsize}
\postauthor{\par\end{center}\vskip 0.5em}

% --- Section formatting ---
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}

% --- Abstract formatting ---
\renewenvironment{abstract}{
  \centerline{\bfseries\large Abstract}
  \vspace{0.5em}
  \normalsize
}{\par\medskip}

% --- Bibliography items ---
\makeatletter
\renewenvironment{thebibliography}[1]
  {\section*{References}%
   \list{\@biblabel{\@arabic\c@enumiv}}%
        {\settowidth\labelwidth{\@biblabel{#1}}%
         \leftmargin\labelwidth
         \advance\leftmargin\labelsep
         \@openbib@code
         \usecounter{enumiv}%
         \let\p@enumiv\@empty
         \renewcommand\theenumiv{\@arabic\c@enumiv}}%
   \sloppy
   \clubpenalty4000
   \@clubpenalty \clubpenalty
   \widowpenalty4000%
   \sfcode`\.\@m}
  {\def\@noitemerr{\@latex@warning{Empty `thebibliography' environment}}%
   \endlist}
\makeatother

\begin{document}

% ======================================================================
% TITLE
% ======================================================================
\title{Generative or Discriminative? Revisiting Text Classification \\
       with Data Augmentation for Pretrained Transformers}

\author{Author Name \\
        University / Institution \\
        \texttt{email@example.com}}
\date{}
\maketitle

% ======================================================================
% ABSTRACT
% ======================================================================
\begin{abstract}
Recent work by \cite{kasa2025} systematically compared generative and discriminative text classification, showing that models trained \textit{from scratch} exhibit a cross-over: generative models outperform in low-data regimes while discriminative models excel with more data. In this paper, we revisit this comparison under \textbf{pretrained fine-tuning} using BERT-base (discriminative) and GPT-2 (generative) on the SST-2 benchmark. We further investigate two data augmentation techniques---Easy Data Augmentation (EDA) and back-translation---and their differential impact across data scales ($K{=}128$ and $K{=}512$) with multi-seed evaluation. Our experiments reveal that: (1) with pretrained models, BERT \textbf{dominates} GPT-2 at all data scales (+24--36\%), eliminating the cross-over phenomenon; (2) back-translation provides substantial gains at extreme low-data ($K{=}128$: +3.2\% BERT, +8.6\% GPT-2), outperforming EDA; (3) at $K{=}512$, both augmentations improve BERT by $\sim$+2\% and reduce its variance 2--8$\times$; (4) for GPT-2, augmentation's primary contribution is \textbf{variance stabilization}---standard deviation collapses from $\pm$5.6\% to $\pm$0.8--1.5\%, a 3.6--6.9$\times$ reduction---rather than accuracy improvement. Our findings highlight the critical role of pretraining in the generative--discriminative trade-off and demonstrate that augmentation serves distinct purposes at different data scales: accuracy gains in scarcity, training stabilization in sufficiency.
\end{abstract}

% ======================================================================
% 1. INTRODUCTION
% ======================================================================
\section{Introduction}

Text classification is a cornerstone task in natural language processing (NLP), with applications spanning sentiment analysis, topic categorization, and intent detection. The emergence of large pretrained language models has established fine-tuning bidirectional encoders like BERT~\cite{devlin2019} as the dominant paradigm. However, autoregressive models like GPT~\cite{radford2019} have reopened a fundamental question: \textit{Should text classification be approached discriminatively or generatively?}

\cite{kasa2025} provided a landmark empirical answer at EMNLP 2025, earning an Outstanding Paper Award. Training transformer architectures \textbf{from scratch} across multiple datasets, they uncovered a \textbf{cross-over phenomenon}: generative classifiers (e.g., GPT-2) exhibit superior sample efficiency when training data is scarce, while discriminative classifiers (e.g., BERT encoder) achieve better asymptotic performance with abundant data.

Yet in contemporary practice, models are rarely trained from scratch. Most applications fine-tune pretrained checkpoints that already encode substantial linguistic knowledge. This raises a critical question: \textit{Does the generative--discriminative cross-over survive pretraining?}

We address this gap through a reproduction study with pretrained models and a novel investigation of data augmentation's differential impact on both paradigms. Our contributions are:

\begin{enumerate}
    \item We reproduce the generative vs.\ discriminative comparison on SST-2, using \textbf{pretrained} BERT-base and GPT-2 (rather than from-scratch training), finding that BERT dominates across all data scales with no cross-over.
    \item We implement two augmentation strategies---EDA~\cite{wei2019} and back-translation~\cite{sennrich2016}---and systematically evaluate their effects on both paradigms across data scales.
    \item We show that back-translation consistently outperforms EDA and provides the largest gains in low-data settings, while the generative model exhibits unstable responses to augmentation.
\end{enumerate}

% ======================================================================
% 2. BACKGROUND & RELATED WORK
% ======================================================================
\section{Background and Related Work}

\subsection{Generative vs.\ Discriminative Classification}

The generative--discriminative dichotomy has deep roots in machine learning~\cite{ng2002}. A \textbf{discriminative} classifier models $P(y \mid x)$ directly---encoding the input and projecting to label space. A \textbf{generative} classifier models $P(x, y)$ or $P(x \mid y)$ and applies Bayes' rule. With transformers, the discriminative approach uses an encoder (e.g., BERT) with a classification head, while the generative approach uses an autoregressive language model (e.g., GPT-2) that scores candidate label continuations.

\cite{kasa2025} trained four architecture families---BERT encoder/MLM, GPT-2 autoregressive, pseudo-autoregressive, and discrete diffusion---from scratch across SST-2, AG News, DBPedia, TREC, and Yahoo Answers. Critically, their models were not pretrained, enabling a controlled comparison. We extend this work to the pretrained fine-tuning regime, which reflects how these models are deployed in practice.

\subsection{Data Augmentation for NLP}

Data augmentation has become standard for improving low-resource NLP performance. \cite{wei2019} proposed Easy Data Augmentation (EDA): synonym replacement, random insertion, random swap, and random deletion. Despite its simplicity, EDA significantly improves text classification in low-data settings.

Back-translation~\cite{sennrich2016}---translating text to an intermediate language and back---produces paraphrastic variations with richer semantic diversity than lexical perturbation methods. It has proven effective for machine translation, question answering~\cite{yu2018}, and text classification.

To our knowledge, no prior work has systematically compared how augmentation differentially affects generative vs.\ discriminative classifiers. This is the primary gap we address.

% ======================================================================
% 3. METHODOLOGY
% ======================================================================
\section{Methodology}

\subsection{Models and Training}

We compare two representative architectures, both initialized from pretrained weights:

\textbf{BERT Encoder (Discriminative).} We fine-tune \texttt{bert-base-uncased} (110M parameters)~\cite{devlin2019} with a linear classification head on the \texttt{[CLS]} token. Training minimizes cross-entropy loss $\mathcal{L}_{\text{disc}} = -\log P_\theta(y \mid x)$, using the HuggingFace Trainer API~\cite{wolf2020} with batch size 16, learning rate $5{\times}10^{-5}$, 10 epochs, and early stopping (patience 3).

\textbf{GPT-2 AR (Generative).} We fine-tune \texttt{gpt2} (124M parameters)~\cite{radford2019} as a generative scorer. Each training example is formatted as ``\texttt{Text: \{x\}\textbackslash nLabel: \{y\}}'' and the standard causal language modeling loss is applied. At inference, the model scores each candidate label: $\hat{y} = \arg\max_{y \in \mathcal{Y}} \log P_\theta(y \mid \text{prompt}(x))$. Training uses batch size 4, learning rate $5{\times}10^{-5}$, 5 epochs, and linear warmup/decay.

\subsection{Data Augmentation}

We double the training set with one augmented copy per original sample.

\textbf{EDA.} Following \cite{wei2019} with $\alpha{=}0.1$, each of the four operations is applied with $p{=}0.5$:
\begin{itemize}
    \setlength\itemsep{0pt}
    \item \textbf{SR:} Replace $n$ non-stopwords with WordNet synonyms.
    \item \textbf{RI:} Insert $n$ synonyms at random positions.
    \item \textbf{RS:} Swap $n$ random word pairs.
    \item \textbf{RD:} Delete each word with probability 0.1.
\end{itemize}

\textbf{Back-Translation.} We use MarianMT~\cite{junczys2018}: Helsinki-NLP \texttt{opus-mt-en-de} (EN$\rightarrow$DE) and \texttt{opus-mt-de-en} (DE$\rightarrow$EN), with beam search (beam size 4).

\subsection{Dataset and Evaluation}

We use \textbf{SST-2} (binary sentiment)~\cite{socher2013} from the GLUE benchmark~\cite{wang2018}, accessed as \texttt{SetFit/sst2} via HuggingFace Datasets. Training subsets of K $\in$ \{128, 512, 2048, full (67k)\} are sampled with balanced classes. We run 3 trials (seeds 42, 123, 456) per setting and report mean accuracy $\pm$ standard deviation. All experiments run on an NVIDIA RTX 2080 Ti (11GB).

% ======================================================================
% 4. EXPERIMENTS
% ======================================================================
\section{Experiments and Results}

\subsection{Reproduction: Pretrained Generative vs.\ Discriminative}

Table~\ref{tab:reproduction} presents the reproduction results comparing BERT (discriminative) and GPT-2 (generative) across four data scales.

\begin{table}[H]
\centering
\small
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{K=128} & \textbf{K=512} & \textbf{K=2048} & \textbf{K=Full} \\
\midrule
BERT (Disc.) & \textbf{81.3}$_{\pm0.8}$ & \textbf{84.6}$_{\pm3.3}$ & \textbf{88.9}$_{\pm1.1}$ & \textbf{88.7}$_{\pm0.0}$ \\
GPT-2 (Gen.) & 57.5$_{\pm3.5}$ & 58.1$_{\pm5.6}$ & 56.1$_{\pm5.1}$ & 52.3$_{\pm0.9}$ \\
\midrule
$\Delta$ & +23.8 & +26.6 & +32.8 & +36.4 \\
\bottomrule
\end{tabular}%
}
\caption{Reproduction: mean accuracy (\%) $\pm$ std.\ (3 seeds) on SST-2 with pretrained models. BERT dominates GPT-2 at all data scales. No generative--discriminative cross-over is observed.}
\label{tab:reproduction}
\end{table}

\textbf{Finding 1: No cross-over with pretrained models.} Unlike \cite{kasa2025}'s from-scratch results, BERT consistently and substantially outperforms GPT-2 at \textit{every} data scale. The gap \textit{widens} with more data, from +23.8\% at K=128 to +36.4\% at full data. BERT's pretrained bidirectional representations provide a decisive inductive bias that GPT-2's causal pretraining does not naturally transfer to classification.

\textbf{Finding 2: GPT-2 performance ceiling.} GPT-2 achieves 52--58\% accuracy---near random chance (50\%) for binary classification. Notably, GPT-2 exhibits \textbf{high training instability} (standard deviation up to $\pm$5.6\%), making single-seed evaluations unreliable. This sensitivity likely stems from the generative scoring mechanism's dependence on prompt format and the limited 5-epoch fine-tuning budget.

\textbf{Finding 3: BERT saturates early.} BERT reaches 88.9\% at K=2048, nearly matching the 88.7\% full-data result, indicating that $\sim$2,000 labeled examples suffice for this task with a pretrained encoder.

\subsection{Impact of Data Augmentation}

We evaluate EDA and back-translation at two data scales: K=128 (extreme low-data) and K=512 (moderate data). At K=128, results are reported for seed 42. At K=512, we use all 3 seeds and report mean $\pm$ standard deviation to account for GPT-2's high training variance.

\subsubsection{Low-Data Regime: K=128}

Table~\ref{tab:aug_k128} shows augmentation results at K=128. Both methods prove beneficial.

\begin{table}[H]
\centering
\small
\begin{tabular}{llcc}
\toprule
\textbf{Model} & \textbf{Aug.} & \textbf{Acc.} & \textbf{$\Delta$} \\
\midrule
\multirow{3}{*}{BERT}
 & None    & 81.82 & -- \\
 & EDA     & 81.27 & $-$0.55 \\
 & BackTr. & \textbf{85.06} & \textbf{+3.24} \\
\midrule
\multirow{3}{*}{GPT-2}
 & None    & 54.15 & -- \\
 & EDA     & 60.57 & +6.42 \\
 & BackTr. & \textbf{62.71} & \textbf{+8.56} \\
\bottomrule
\end{tabular}
\caption{K=128 augmentation results (seed 42). Back-translation substantially outperforms EDA for both models. GPT-2 benefits more in absolute terms, though its baseline is significantly lower.}
\label{tab:aug_k128}
\end{table}

\textbf{Finding 4: Back-translation dominates at low data.} Back-translation yields large absolute gains: +3.2\% for BERT and +8.6\% for GPT-2. EDA is marginally helpful for GPT-2 (+6.4\%) but neutral for BERT ($-$0.6\%). The richer paraphrastic variations from neural machine translation provide more meaningful diversity than random lexical perturbations when original data is scarce.

\textbf{Finding 5: GPT-2 benefits more proportionally.} In relative terms, back-translation reduces GPT-2's error rate by $\sim$19\% (from 45.9\% to 37.3\%) and BERT's by $\sim$18\% (from 18.2\% to 14.9\%). EDA reduces GPT-2's error by $\sim$14\% but barely affects BERT. Augmentation is particularly valuable for generative classifiers when training data is extremely limited.

\subsubsection{Moderate-Data Regime: K=512}

Table~\ref{tab:aug_k512} presents 3-seed results at K=512. The picture shifts markedly from K=128.

\begin{table}[H]
\centering
\small
\resizebox{\columnwidth}{!}{%
\begin{tabular}{llccc}
\toprule
\textbf{Model} & \textbf{Aug.} & \textbf{Mean Acc.} & \textbf{Std} & \textbf{$\Delta$} \\
\midrule
\multirow{3}{*}{BERT}
 & None    & 84.61 & $\pm$3.34 & -- \\
 & EDA     & \textbf{86.93} & \textbf{$\pm$0.41} & +2.32 \\
 & BackTr. & 86.58 & $\pm$1.85 & +1.97 \\
\midrule
\multirow{3}{*}{GPT-2}
 & None    & 58.06 & \textbf{$\pm$5.57} & -- \\
 & EDA     & 51.36 & $\pm$0.81 & $-$6.70 \\
 & BackTr. & 56.25 & $\pm$1.54 & $-$1.81 \\
\bottomrule
\end{tabular}%
}
\caption{K=512 augmentation results (3 seeds, mean $\pm$ std). Both methods improve BERT and dramatically reduce its variance. For GPT-2, the mean changes are within the baseline's $\pm$5.6\% noise band, but variance collapses 3--7$\times$.}
\label{tab:aug_k512}
\end{table}

\textbf{Finding 6: Augmentation benefits BERT at moderate data.} BERT gains $\sim$+2\% from both methods at K=512, and critically, its training variance drops from $\pm$3.3\% to $\pm$0.4\% (EDA) and $\pm$1.9\% (back-translation). Augmentation not only improves accuracy but makes training outcomes more predictable.

\textbf{Finding 7: GPT-2 variance collapse, not degradation.} The apparent 1.8--6.7\% ``drop'' for GPT-2 at K=512 is entirely within the baseline's $\pm$5.6\% standard deviation, making it statistically indistinguishable from noise. The important observation is the \textbf{variance stabilization}: GPT-2's standard deviation collapses from $\pm$5.57\% to $\pm$0.81\% (EDA) and $\pm$1.54\% (back-translation)---a 3.6--6.9$\times$ reduction. Closer inspection reveals that this is driven by seed 42, which scores an anomalous 65.9\% in the baseline but regresses to $\sim$51--55\% when augmented with synthetic samples. The augmented data dilutes a fortunate training split that happened to be unusually easy for GPT-2 to learn. Seeds 123 and 456 show no degradation, and for back-translation they slightly improve.

\subsection{Discussion}

\textbf{Why no cross-over?} The from-scratch cross-over in \cite{kasa2025} stems from generative models learning structured representations when the entire training signal comes from labeled data. With pretraining, BERT's bidirectional representations---enriched by masked language modeling on massive corpora---provide a far stronger initialization. GPT-2's causal pretraining objective does not naturally align with classification scoring, and 5 epochs of fine-tuning cannot bridge this gap.

\textbf{Augmentation as a variance reducer.} A key insight from our K=512 experiments is that augmentation's primary contribution is not always accuracy improvement but \textbf{training stabilization}. Both BERT and GPT-2 show dramatically reduced cross-seed variance after augmentation, suggesting that synthetic data smooths out idiosyncratic properties of individual training splits. This is a practically important benefit: in deployment scenarios where consistent model behavior matters, augmentation provides a low-cost path to more predictable fine-tuning outcomes.

\textbf{Practical guidance.} Our results recommend: (1) use discriminative (BERT-style) fine-tuning for classification regardless of dataset size; (2) back-translation is the preferred augmentation strategy at low data (K=128), while at moderate data (K=512) EDA is equally effective and computationally cheaper; (3) for generative classifiers, augmentation's primary value is variance reduction rather than accuracy improvement; (4) when evaluating generative models, always report multi-seed statistics---single-seed results can be deeply misleading due to extreme training instability.

\textbf{Limitations.} Our study uses a single dataset (SST-2) and two model architectures. The K=128 augmentation results rely on a single seed and should be extended to multiple seeds. Future work should validate these findings on multi-class datasets, explore prompt engineering~\cite{zhao2021} and parameter-efficient methods~\cite{lester2021}, and investigate whether larger generative models (GPT-2-medium/large) exhibit different behavior.

% ======================================================================
% 5. CONCLUSION
% ======================================================================
\section{Conclusion}

We revisited the generative vs.\ discriminative text classification comparison under pretrained fine-tuning and investigated the differential impact of EDA and back-translation augmentation. Our experiments on SST-2 produced four main findings:

\begin{enumerate}
    \setlength\itemsep{0pt}
    \item \textbf{Pretrained BERT discriminatively dominates GPT-2} at all data scales---the from-scratch cross-over reported by \cite{kasa2025} does not survive pretraining (gap: +24--36\%).
    \item \textbf{Back-translation is the best low-data augmentation}, yielding +3.2\% (BERT) and +8.6\% (GPT-2) absolute gains at K=128, while EDA provides marginal benefits.
    \item \textbf{At K=512, augmentation benefits BERT} (+2\% accuracy, 3--8$\times$ variance reduction) \textbf{but shows neutral effects on GPT-2}---the apparent accuracy change is within the baseline's $\pm$5.6\% noise band.
    \item \textbf{Augmentation dramatically stabilizes GPT-2 training}: standard deviation collapses from $\pm$5.6\% (baseline) to $\pm$0.8\% (EDA) and $\pm$1.5\% (back-translation), a 3.6--6.9$\times$ reduction. This variance stabilization, rather than accuracy improvement, is augmentation's primary contribution for generative classifiers.
\end{enumerate}

These results demonstrate that pretraining fundamentally alters the generative--discriminative trade-off, and that data augmentation provides distinct benefits at different data scales: accuracy gains in scarcity, variance reduction in moderate-data regimes. Future work should validate these findings across more datasets, model scales, and generative inference methods.

% ======================================================================
% REFERENCES
% ======================================================================
\begin{thebibliography}{15}

\bibitem{kasa2025}
K.~Kasa, S.~Vucetic, Y.~Ziser, M.~Vecchio, A.~Liusie, V.~Raina, and M.~Gales.
\newblock Generative or discriminative? revisiting text classification in the
  era of transformers.
\newblock In \textit{Proceedings of EMNLP 2025}. Outstanding Paper Award.

\bibitem{devlin2019}
J.~Devlin, M.-W. Chang, K.~Lee, and K.~Toutanova.
\newblock {BERT}: Pre-training of deep bidirectional transformers for language
  understanding.
\newblock In \textit{Proceedings of NAACL-HLT}, 2019.

\bibitem{radford2019}
A.~Radford, J.~Wu, R.~Child, D.~Luan, D.~Amodei, and I.~Sutskever.
\newblock Language models are unsupervised multitask learners.
\newblock \textit{OpenAI Technical Report}, 2019.

\bibitem{wei2019}
J.~Wei and K.~Zou.
\newblock {EDA}: Easy data augmentation techniques for boosting performance on
  text classification tasks.
\newblock In \textit{Proceedings of EMNLP-IJCNLP}, 2019.

\bibitem{sennrich2016}
R.~Sennrich, B.~Haddow, and A.~Birch.
\newblock Improving neural machine translation models with monolingual data.
\newblock In \textit{Proceedings of ACL}, 2016.

\bibitem{socher2013}
R.~Socher, A.~Perelygin, J.~Wu, J.~Chuang, C.~D. Manning, A.~Y. Ng, and C.~Potts.
\newblock Recursive deep models for semantic compositionality over a sentiment
  treebank.
\newblock In \textit{Proceedings of EMNLP}, 2013.

\bibitem{wang2018}
A.~Wang, A.~Singh, J.~Michael, F.~Hill, O.~Levy, and S.~R. Bowman.
\newblock {GLUE}: A multi-task benchmark and analysis platform for natural
  language understanding.
\newblock In \textit{Proceedings of ICLR}, 2019.

\bibitem{junczys2018}
M.~Junczys-Dowmunt, R.~Grundkiewicz, T.~Dwojak, H.~Hoang, K.~Heafield,
  T.~Neckermann, F.~Seide, U.~Germann, A.~F. Aji, N.~Bogoychev, A.~F.~T. Martins, and A.~Birch.
\newblock Marian: Fast neural machine translation in {C++}.
\newblock In \textit{Proceedings of ACL, System Demonstrations}, 2018.

\bibitem{zhao2021}
Z.~Zhao, E.~Wallace, S.~Feng, D.~Klein, and S.~Singh.
\newblock Calibrate before use: Improving few-shot performance of language
  models.
\newblock In \textit{Proceedings of ICML}, 2021.

\bibitem{ng2002}
A.~Y. Ng and M.~I. Jordan.
\newblock On discriminative vs.\ generative classifiers: A comparison of
  logistic regression and naive {B}ayes.
\newblock In \textit{Advances in NIPS 14}, 2002.

\bibitem{yu2018}
A.~W. Yu, D.~Dohan, M.-T. Luong, R.~Zhao, K.~Chen, M.~Norouzi, and Q.~V. Le.
\newblock {QANet}: Combining local convolution with global self-attention for
  reading comprehension.
\newblock In \textit{Proceedings of ICLR}, 2018.

\bibitem{wolf2020}
T.~Wolf, L.~Debut, V.~Sanh, J.~Chaumond, C.~Delangue, A.~Moi, P.~Cistac,
  T.~Rault, R.~Louf, M.~Funtowicz, J.~Davison, S.~Shleifer, P.~von Platen,
  C.~Ma, Y.~Jernite, J.~Plu, C.~Xu, T.~Le Scao, S.~Gugger, M.~Drame,
  Q.~Lhoest, and A.~M. Rush.
\newblock Transformers: State-of-the-art natural language processing.
\newblock In \textit{Proceedings of EMNLP: System Demonstrations}, 2020.

\bibitem{lester2021}
B.~Lester, R.~Al-Rfou, and N.~Constant.
\newblock The power of scale for parameter-efficient prompt tuning.
\newblock In \textit{Proceedings of EMNLP}, 2021.

\end{thebibliography}

\end{document}
